---
title: "Generate_markerDB"
author: "Yuna Blum"
date: "`r Sys.Date()`"
output:  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generate_markerDB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Literature marker databases

Two different literature marker databases were downloaded: 

* CellMarker  (Zhang et al., 2019, DOI: 10.1093/nar/gky900).  
Link to dowload the database: http://bio-bigdata.hrbmu.edu.cn/CellMarker/download.jsp
We choose the human restricted marker file: Human_cell_markers.txt  



* CEllmatch (Shao et al, iScience 2020, https://doi.org/10.1016/j.isci.2020.100882), which is included in the scCATCH tool.  
Link to download the database: https://github.com/ZJUFanLab/scCATCH/tree/master/R  
"CellMatch reference database was derived from various resources, such as CellMarker (Zhang et al., 2019b), MCA (Han et al., 2018), CancerSEA (Yuan et al., 2019), and the CD Marker Handbook."


```{r download}
#CellMarker
cellmarker = read.table(url("http://biocc.hrbmu.edu.cn/CellMarker/download/Human_cell_markers.txt"), header=TRUE, sep="\t")

#CEllmatch
sysdata_cellmatch=tempfile()
curl::curl_download(url="https://github.com/ZJUFanLab/scCATCH/raw/master/data/cellmatch.rda", sysdata_cellmatch)
load(sysdata_cellmatch)
# cellmatch = CellMatch
unlink(sysdata_cellmatch)
ls()

```

## Filters on the databases

Some filters are applied on the databases:
* Restrict markers to human markers if several species are included.
* Restrict to available gene symbols.
* Set a minimal number of genes per cell population 
* Set a maximum number of genes per cell population 

```{r filter cellmarker}

### Filtering the cellmarker object

## Remove NA values
cellmarker = cellmarker[-which(is.na(cellmarker$geneSymbol)), ]
head(cellmarker)

# Retrieve all the genes per cell population: each line corresponds to a particular cell populations. The gene markers are separated by a coma.
allgenes = lapply(sapply(cellmarker$geneSymbol,function(x){strsplit(as.character(x),", ")}), function(x){genes = unique(gsub("\\[|]", "", x)) ; return(genes)})

# Prepare the names of cell population mentionning the tissue type and cell type
cellID = paste(gsub(" ", "_", cellmarker$tissueType),  gsub(" ", "_", cellmarker$cellType), gsub("[+]", "plus", gsub("[(]|[)]|-","", gsub(" ", "_", cellmarker$cellName))), sep=".")

# Transform the df into a list. Note that some cell populations are specified several times in the file, with common genes. In addition we remove the NA values.
liter_cellmarker = tapply(cellmarker$geneSymbol, cellID, function(x){genes = unique(unlist(strsplit(as.character(x),", "))) ; if(sum(genes=="NA")>0){genes = genes[-which(genes=="NA")]} ; return(genes)})

## Filtering for the number of genes.

nbgenes = unlist(lapply(liter_cellmarker, length))
liter_cellmarker_filter = liter_cellmarker[which(nbgenes>=3 & nbgenes<=200)]

length(liter_cellmarker_filter)
liter_cellmarker_filter[1:3]

```

```{r filter cellmatch}

### Filtering the cellmarker object

## Keep only human markers
cellmatch = cellmatch[cellmatch$species == "Human",]
head(cellmatch)

## Prepare the names of cell population mentionning the tissue type and cell type
cellID = paste(gsub(" ", "_", cellmatch$tissue),  gsub(" ", "_", cellmatch$celltype), gsub("[+]", "plus", gsub("[(]|[)]|-","", gsub(" ", "_", cellmatch$cellName))), sep=".")

## Transform the df into a list. 
liter_cellmatch = tapply(cellmatch$gene, cellID, list)
liter_cellmatch = lapply(liter_cellmatch, function(x) unique(x))

## Filtering for the number of genes.
nbgenes = unlist(lapply(liter_cellmatch, length))
liter_cellmatch_filter = liter_cellmatch[which(nbgenes>=3 & nbgenes<=200)]

length(liter_cellmatch_filter)
liter_cellmatch_filter[1:3]

```
## Saving databases

```{r save, eval=FALSE}
saveRDS(liter_cellmarker_filter, "cellmarkerDB.RDS")

saveRDS(liter_cellmatch_filter, "cellmatchDB.RDS")

```

Note (i) that the last code is not running when running the vignette. 
Note (ii) that the marker DB objects are in the list format as it is required in the fgsea function from the R package fgsea.


